---
title: "Lecture 2: Introduction to R"
format: html
author: "Rozanne Lim" 
date: 08/26/25
---

# Installing and loading R packages 

```{r}
install.packages()
```

To see what other packages are required by ggplot2 
```{r}
tools::package_dependencies("ggplot2")
```

Use the remotes package to install packages from GitHub
```{r}
remotes::install_github()
```

Once packages are installed, they do not have to be re-installed. They do, however, need to be loaded in to each new R session.

```{r}
library() 
```


## cloning repository from github

```{r}
git clone https://github.com/YOUR-USERNAME/YOUR-REPOSITORY.git
```


## in terminal: repo new working directory, list the files, read the README file, show the remote URL
cd myrepo
ls
head README.md
git remote show origin

## in terminal: add a line to the README file and check the status of the repository
echo "A line I wrote on my local computer  " >> README.md
git status

##rewriting credentials
gitcreds::gitcreds_set()

##code for creating a new R project and pushing it to github
if (!requireNamespace("usethis", quietly = TRUE)) {
  install.packages("usethis")
}

usethis::create_project("~/Desktop/biostat776classnotes")

## Then in your new RStudio project, run the following:
usethis::use_git()
usethis::use_github()

##reproducable code example
install.packages("reprex")
reprex::reprex()

