---
title: "Lecture 13: Control Structures"
format: html
author: "Rozanne Lim"
---

**Commonly used control structures are**
- if and else: testing a condition and acting on it
- for: execute a loop a fixed number of times
- while: execute a loop while a condition is true
- repeat: execute an infinite loop (must break out of it to stop)
- break: break the execution of a loop
- next: skip an interation of a loop

**if else example**
```{r}
if (<condition1>) {
    ## do something
} else if (<condition2>)  {
    ## do something different
} else {
    ## do something entirely different
}
```

**another way of writing**
```{r}
y <- if (x > 3) {
    10
} else {
    0
}

y
```

**Practice Question**
Let’s use the palmerpenguins dataset and write a if-else statement that
1. Randomly samples a value from a standard normal distribution (Hint: check out the rnorm(n, mean = 0, sd = 1) function in base R).
2. If the value is larger than 0, use dplyr functions to keep only the Chinstrap penguins.
3. Otherwise, keep only the Gentoo penguins.
4. Re-run the code 10 times and look at the output.

```{r}
library(palmerpenguins)
library(tidyverse)
penguins <- penguins

x <- rnorm(1, mean = 0, sd = 1)
if (x > 0){
    penguins %>%
        filter(species == "Chinstrap") 
} else {
    penguins %>%
        filter(species == "Gentoo")
}

```