---
title: "RegEx"
format: html
embed-resources: TRUE
---

A regular expression (also known as a “regex” or “regexp”) is a concise language for describing patterns in character strings.

RegEx can be used for:
1. Searching for a pattern or string within another string (e.g searching for the string “a” in the string “Maryland”)
2. Replacing one part of a string with another string (e.g replacing the string “t” with “p” in the string “hot” where you are changing the string “hot” to “hop”)

## repetition

You can specify a regular expression that contains a certain number of characters or metacharacters using the enumeration metacharacters (or sometimes called quantifiers).

+: indicates that one or more of the preceding expression should be present (or matches at least 1 time)
*: indicates that zero or more of the preceding expression is present (or matches at least 0 times)
?: indicates that zero or 1 of the preceding expression is not present or present at most 1 time (or matches between 0 and 1 times)

Let’s practice a few out together. Make the following regular expressions for the character string “spookyhalloween”:
1. Does “zz” appear 1 or more times?
2. Does “ee” appear 1 or more times?
3. Does “oo” appear 0 or more times?
4. Does “ii” appear 0 or more times?

```{r}
grepl("(zz)+", "spookyhalloween")
grepl("(ee)+", "spookyhalloween")
grepl("(oo)*", "spookyhalloween")
grepl("(ii)*", "spookyhalloween")
```

**Specify exact numbers of expressions**
{n}: exactly n
{n,}: n or more
{,m}: at most m
{n,m}: between n and m

Example 1:
```{r}
# Does "Mississippi" contain exactly 2 adjacent "s" ?
grepl("s{2}", "Mississippi")

# This is equivalent to the expression above:
grepl("ss", "Mississippi")
```

Example 2:
```{r}
# Does "Mississippi" contain between 1 and 3 adjacent "s" ?
grepl("s{1,3}", "Mississippi")
```

Example 3: 
```{r}
# Does "Mississippi" contain between 2 and 3 adjacent "i" ?
grepl("i{2,3}", "Mississippi")
```

Let’s practice a few out together. Make the following regular expressions for the character string “spookyspookyhalloweenspookyspookyhalloween”:

Search for “spooky” exactly 2 times. What about 3 times?
Search for “spooky” exactly 2 times followed by any character of length 9 (i.e. “halloween”).
Same search as above, but search for that twice in a row.
Same search as above, but search for that three times in a row.

```{r}
grepl("(spooky){2}", "spookyspookyhalloweenspookyspookyhalloween")
grepl("(spooky){3}", "spookyspookyhalloweenspookyspookyhalloween")
#search for spooky exactly 2 times followeed by any character of length 9
grepl("(spooky){2}.{9}", "spookyspookyhalloweenspookyspookyhalloween")
grepl("((spooky){2}){2}", "spookyspookyhalloweenspookyspookyhalloween")
```